---
title: "Sherlock Holmes Case Using Bayes Net"
author: "Jason Chan"
date: "January 21, 2017"
output: 
  html_document:
    toc: TRUE
---

#Libraries Used
```{r, message=FALSE, error=FALSE, warning=FALSE}
library(gRain)
library(gRbase)
library(ggm)
library(Rgraphviz)
```
  
<br />

#The Case
**Sherlock Holmes was asked to solve the following problem:**

A man was murdered last night an the Police has three suspects. A knife was found close to the body and the expert examination tests showed that it had the fingerprints of suspect 3. 

A neighbor saw a man running out of the house in which the murder occurred and the man had the same height of suspect 1 being much smaller than suspects 2 and 3.

**Which suspect is the murderer with highest probability** ?
  
<br />

#Step 1: Visualization Plot
```{r, echo=FALSE}
g <- list(~criminal, ~scene | criminal , ~knife | criminal , ~expert | knife, ~height | scene)

crimedag <- dagList(g)

plot(crimedag)
```

Where:

* Criminal - The criminal responsible for the murder
* Knife - The last person who handled the knife
* Scene - Person who left the scene 
* Height - The person's height matching that of the suspects
* Expert - Expert's examination of which suspect's fingerprint found
  
<br />

#Step 2: Assigning the probabilities

Assuming that we consulted a Crime Scene Investigator for this domain knowledge and this were his proposed probabilities:

```{r, echo=TRUE}
#Specify the levels
suspect <- c("one", "two", "three")

#Construct Conditional Probability Table (CPT)
c <- cptable(~criminal, values = c(1/3, 1/3, 1/3), levels = suspect)

s.c <- cptable(~scene | criminal, values = c(0.8, 0.1, 0.1, 0.1, 0.8, 0.1, 0.1, 0.1, 0.8), levels = suspect)

h.s <- cptable(~height | scene, values = c(0.6, 0.3, 0.1, 0.1, 0.6, 0.3, 0.1, 0.4, 0.5), levels = suspect)

k.c <- cptable(~knife | criminal, values = c(0.9, 0.05, 0.05, 0.05, 0.9, 0.05, 0.05, 0.05, 0.9), levels = suspect)

e.k <- cptable(~expert | knife, values = c(0.8, 0.1, 0.1, 0.1, 0.8, 0.1, 0.1, 0.1, 0.8), levels = suspect)

#Compile the Network
plist <- compileCPT(list(c, s.c, h.s, k.c, e.k))
grn1 <- grain(plist)
plot(grn1)


```
  
<br />

#Step 3: Set Evidence and Run Query

Note that from the case briefing above, we can find the evidence from these two extracts:

* A knife was found close to the body and the tests showed that it had the fingerprints of **suspect 3**
* the man had the same height of **suspect 1**

This simple implies setting the evidence of `height = "one"` and `expert = "three"` and running the query.

```{r}
find1 <- setFinding(grn1, nodes = c("height", "expert"), states = c("one", "three"))

querygrain(find1, nodes = c("criminal"), type = "marginal")
```

From the probability above, it seems like we have found our most likely murderer, which is **suspect 3** ! Case solved ! Note that without any evidence, probability of each suspect is a murderer stays at 1/3.
  
<br />

#Other Queries Example

##Example 1
With the evidence as above, what is the probability of that the suspects were at the scene and held the knife ?

```{r}
querygrain(find1, nodes = c("scene", "knife"), type = "joint")
```

Focusing on the diagonal probabilities, it seems like suspect 1 and 3 were almost equal of likelihood to have held the knife and at the scene.

##Example 2
With the evidence as above, what is the probability of the suspects being the criminal given that they were at the scene ?
```{r}
querygrain(grn1, nodes = c("criminal", "scene"), type = "conditional")
```

Basically gets back the values of our Conditional Probability Tables. This is a good way to check entered values.